---
title: "Simulating pediatric body size from NHANES"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating pediatric body size from NHANES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

```{r setup, warning=F, message=F}
library(pmxTools)
```

Pediatric simulations need realistic covariates. Body weight drives allometric
scaling of clearance and volume, and weight-based or body-surface-area-based
dosing needs weight and height that belong together. `sample_nhanes()`
generates virtual children aged 2 to 17 years with body weight and/or
standing height drawn from the US National Health and Nutrition Examination
Survey (NHANES).

## The reference data

The package includes `nhanes_peds`, with one row per child from four NHANES
releases (2013-2014 to 2021-2023):

```{r}
head(nhanes_peds)
table(nhanes_peds$CYCLE)
```

Each child has a mobile examination center (MEC) exam weight (`MEC_WT`), which
NHANES provides so that its sample represents the US population. A few
children have only one of weight or height measured. They are kept, and are
used whenever the missing measure isn't needed.

## Simulating a population

`sample_nhanes()` simulates `n` children (500 by default) per age (in whole
years) and sex:

```{r}
sim <- sample_nhanes(seed = 20261008)
sim
```

For each age and sex, it resamples real children, weighted by their MEC exam
weights with each NHANES release given an equal share. It then adds a small
amount of correlated random noise to log weight and log height. The simulated
values are therefore continuous rather than copies of the reference children,
and a tall child stays heavier than a short one of the same age. `SOURCE_CYCLE`
and `SOURCE_SEQN` identify the real child each simulated child was drawn from.

The amount of smoothing in each age and sex group is stored with the result:

```{r}
head(attr(sim, "kernels"))
```

`N_NHANES` is the number of reference children, `N_EFF` their effective sample
size after weighting, `H_WT` and `H_HT` the noise standard deviations on the
log scale, and `RHO` the weighted correlation between log weight and log
height.

## Checking the simulation

`compare_nhanes()` summarises weight, height and body mass index for the
simulated population and for the weighted NHANES reference it came from:

```{r}
cmp <- compare_nhanes(sim)
cmp[cmp$AGE == 10, c("SEX", "VARIABLE", "Median_REF", "Median_SIM",
                     "P05_RATIO", "P95_RATIO", "RHO_REF", "RHO_SIM")]
```

`plot_nhanes()` shows the 5th, 50th and 95th percentiles against age:

```{r}
plot_nhanes(cmp)
```

Body mass index is included because it depends on weight and height together.
If it matches the reference, the relationship between the two has been kept,
not just each measure on its own. The 5th and 95th percentiles of weight vary
more from run to run than the medians, because weight has a long upper tail;
increase `n` if the tails matter for your simulation.

## Options

Simulate a subset of ages, one sex, or a single measure. With
`vars = "WT"`, children whose height is missing are also used:

```{r}
wt <- sample_nhanes(n = 100, ages = 6:12, sex = "Female", vars = "WT",
                    seed = 1)
head(wt)
```

`cycles` restricts the reference to particular NHANES releases, for example
the most recent one only:

```{r}
recent <- sample_nhanes(n = 100, ages = 2:5, cycles = "2021-23", seed = 1)
table(recent$SOURCE_CYCLE)
```

`bandwidth_factor` scales the smoothing. `0` returns exact copies of real
children's measurements; values above 1 smooth more.

```{r}
copies <- sample_nhanes(n = 5, ages = 8, sex = "Male", bandwidth_factor = 0,
                        seed = 1)
copies
```

## Using the population in a PK simulation

The result is an ordinary data frame, so derived covariates are one step away.
For example, body surface area (Mosteller) and an allometrically scaled
clearance for a typical value of 10 L/h at 70 kg:

```{r}
sim$BSA <- sqrt(sim$WT * sim$HT / 3600)
sim$CL <- 10 * (sim$WT / 70)^0.75
head(sim[c("ID", "AGE", "SEX", "WT", "HT", "BSA", "CL")])
```
