---
title: "Simulating pediatric body size from NHANES"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating pediatric body size from NHANES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6
)
```

```{r setup, warning=F, message=F}
library(pmxTools)
```

Pediatric simulations need realistic covariates. Body weight drives allometric
scaling of clearance and volume, and weight-based or body-surface-area-based
dosing needs weight and height that belong together. `sample_nhanes()`
generates virtual children aged 2 to 17 years with body weight and/or
standing height drawn from the US National Health and Nutrition Examination
Survey (NHANES).

## The reference data

The package includes `nhanes_peds`, with one row per child from four NHANES
releases (2013-2014 to 2021-2023):

```{r}
head(nhanes_peds)
table(nhanes_peds$CYCLE)
```

Each child has a mobile examination center (MEC) exam weight (`MEC_WT`), which
NHANES provides so that its sample represents the US population. A few
children have only one of weight or height measured. They are kept, and are
used whenever the missing measure isn't needed.

## Simulating a population

`sample_nhanes()` simulates `n` children (500 by default) per age (in whole
years) and sex:

```{r}
sim <- sample_nhanes(seed = 20261008)
sim
```

For each age and sex, it resamples real children, weighted by their MEC exam
weights with each NHANES release given an equal share. Resampling is by
child: each simulated child takes the weight and height of one real child, so
every value is one that was actually measured, and weight and height stay
paired. `SOURCE_CYCLE` and `SOURCE_SEQN` identify that child.

Details of the reference in each age and sex group are stored with the result:

```{r}
head(attr(sim, "kernels"))
```

`N_NHANES` is the number of reference children, `N_EFF` their effective sample
size after weighting, and `RHO` the weighted correlation between log weight
and log height. `H_WT` and `H_HT` are 0 here; they are the noise used by the
smoothed method described below.

## Checking the simulation

`compare_nhanes()` summarises weight, height and body mass index for the
simulated population and for the weighted NHANES reference it came from:

```{r}
cmp <- compare_nhanes(sim)
cmp[cmp$AGE == 10, c("SEX", "VARIABLE", "Median_REF", "Median_SIM",
                     "P05_RATIO", "P95_RATIO", "RHO_REF", "RHO_SIM")]
```

`plot_nhanes()` shows the 5th, 50th and 95th percentiles against age:

```{r}
plot_nhanes(cmp)
```

Body mass index is included because it depends on weight and height together.
If it matches the reference, the relationship between the two has been kept,
not just each measure on its own. The 5th and 95th percentiles of weight vary
more from run to run than the medians, because weight has a long upper tail;
increase `n` if the tails matter for your simulation.

## Options

Simulate a subset of ages, one sex, or a single measure. With
`vars = "WT"`, children whose height is missing are also used:

```{r}
wt <- sample_nhanes(n = 100, ages = 6:12, sex = "Female", vars = "WT",
                    seed = 1)
head(wt)
```

`cycles` restricts the reference to particular NHANES releases, for example
the most recent one only:

```{r}
recent <- sample_nhanes(n = 100, ages = 2:5, cycles = "2021-23", seed = 1)
table(recent$SOURCE_CYCLE)
```

## Smoothed values

Resampled values repeat those of the reference children. If a simulation
needs continuous values instead, `method = "smooth"` adds a small amount of
random noise to each resampled child's log weight and log height. The noise
for weight and height is correlated like the reference data, so a tall child
stays heavier than a short one of the same age:

```{r}
smoothed <- sample_nhanes(n = 500, ages = c(4, 10, 16), method = "smooth",
                          seed = 20261008)
head(smoothed)
attr(smoothed, "kernels")
```

`H_WT` and `H_HT` are the noise standard deviations on the log scale (0.05
is roughly 5%). `bandwidth_factor` scales them: values above 1 smooth more.
Check smoothed populations with `compare_nhanes()` in the same way.

## Using the population in a PK simulation

The result is an ordinary data frame, so derived covariates are one step away.
For example, body surface area (Mosteller) and an allometrically scaled
clearance for a typical value of 10 L/h at 70 kg:

```{r}
sim$BSA <- sqrt(sim$WT * sim$HT / 3600)
sim$CL <- 10 * (sim$WT / 70)^0.75
head(sim[c("ID", "AGE", "SEX", "WT", "HT", "BSA", "CL")])
```
