---
title: "Drawing PK curves with pmxTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drawing PK curves with pmxTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=F, message=F}
library(pmxTools)
library(gridExtra)
```

`pmxTools` includes functions for constructing pharmacokinetic (PK) concentration-time curves for a range of common compartmental systems using PK parameters and microconstants. The underlying solutions are based on Julie Bertrand and France Mentré's [Mathematical Expressions of the Pharmacokinetic and Pharmacodynamic Models implemented in the Monolix software](http://lixoft.com/wp-content/uploads/2016/03/PKPDlibrary.pdf), published in 2008. 

The options available are by necessity confined to one-, two- and three-compartmental linear models with intravenous bolus, infusion or oral dosing (the latter with zero- or first-order absorption). If more complex systems are needed, you are probably better off using an ODE-based simulation package such as [`RxODE`](https://nlmixrdevelopment.github.io/RxODE/) or [`mrgSolve`](https://mrgsolve.org/). 

## Specifying model parameters

Models take dose, time and PK parameters as inputs. Names for some PK parameters, especially volumes and intercompartmental clearances, vary by model type. 

Parameter | Description                                        | Used for 
----------|----------------------------------------------------|------------------------------------------------
`t`         | Time                                               | Single-dose 
`tad`       | Time after last dose                               | Steady state 
`dose`      | Dose                                               | All 
`CL`        | Clearance                                          | All 
`V`, `V1`     | Central volume of distribution                     | All 
`V2`        | First peripheral volume of distribution            | Two- and three-compartment models
`V3`        | Second peripheral volume of distribution           | Three-compartment models
`Q`, `Q1`, `Q2` | Intercompartmental clearance (between V1 and V2)   | Two- and three-compartment models
`Q3`        | Intercompartmental clearance (between V1 and V3)   | Three-compartment models
`tlag`      | Lag time                                           | Models with lag time
`tinf`      | Duration of infusion                               | Models with infusion
`dur`       | Duration of zero-order absorption                  | Models with zero-order absorption
`ka`        | First-order absorption rate                        | Models with first-order absorption


## One-compartment, single dose 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
1-compartment   | Single-dose, bolus                                     | `calc_sd_1cmt_linear_bolus(t, dose, CL, V)`
&nbsp;          | Single-dose, infusion                                  | `calc_sd_1cmt_linear_infusion(t, dose, CL, V, tinf)` 
&nbsp;          | Single-dose, zero-order oral absorption                | `calc_sd_1cmt_linear_oral_0(t, dose, CL, V, dur)`
&nbsp;          | Single-dose, zero-order oral absorption with lag time  | `calc_sd_1cmt_linear_oral_0_lag(t, dose, CL, V, dur, tlag)`
&nbsp;          | Single-dose, first-order oral absorption               | `calc_sd_1cmt_linear_oral_1(t, dose, CL, V, ka)` 
&nbsp;          | Single-dose, first-order oral absorption with lag time | `calc_sd_1cmt_linear_oral_1_lag(t, dose, CL, V, ka, tlag)` 

For example -

```{r 1cptdemo, fig.height=10, fig.width=7,message=FALSE}

library(ggplot2)

t <- seq(0, 24, by=0.1)

df1 <- data.frame(t = t,
                  Cp1 = calc_sd_1cmt_linear_bolus(t = t, dose = 600, CL = 5, V = 20),
                  Cp2 = calc_sd_1cmt_linear_infusion(t = t, dose = 600, CL = 5, V = 20, tinf=2),
                  Cp3 = calc_sd_1cmt_linear_oral_0(t = t, dose = 600, CL = 5, V = 20, dur=2),
                  Cp4 = calc_sd_1cmt_linear_oral_0_lag(t = t, dose = 600, CL = 5, V = 20, dur=2, tlag=1),
                  Cp5 = calc_sd_1cmt_linear_oral_1(t = t, dose = 600, CL = 5, V = 20, ka=1.5),
                  Cp6 = calc_sd_1cmt_linear_oral_1_lag(t = t, dose = 600, CL = 5, V = 20, ka=1.5, tlag=1))

p1.1 <- ggplot(df1, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p1.2 <- ggplot(df1, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p1.3 <- ggplot(df1, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p1.4 <- ggplot(df1, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p1.5 <- ggplot(df1, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p1.6 <- ggplot(df1, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p1.1, p1.2, p1.3, p1.4, p1.5, p1.6, nrow=3)

```

## One-compartment, steady state 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
1-compartment   | Steady state, bolus                                     | `calc_ss_1cmt_linear_bolus(tad, tau, dose, CL, V)`
&nbsp;          | Steady state, infusion                                  | `calc_ss_1cmt_linear_infusion(tad, tau, dose, CL, V, tinf)` 
&nbsp;          | Steady state, zero-order oral absorption                | `calc_ss_1cmt_linear_oral_0(tad, tau, dose, CL, V, dur)`
&nbsp;          | Steady state, zero-order oral absorption with lag time  | `calc_ss_1cmt_linear_oral_0_lag(tad, tau, dose, CL, V, dur, tlag)`
&nbsp;          | Steady state, first-order oral absorption               | `calc_ss_1cmt_linear_oral_1(tad, tau, dose, CL, V, ka)` 
&nbsp;          | Steady state, first-order oral absorption with lag time | `calc_ss_1cmt_linear_oral_1_lag(tad, tau, dose, CL, V, ka, tlag)` 

For example -

```{r 1cptssdemo, fig.height=10, fig.width=7,message=FALSE}

t <- seq(0, 24, by=0.1)

df1ss <- data.frame(t = t,
                  Cp1 = calc_ss_1cmt_linear_bolus(t = t, tau = 24, dose = 600, CL = 1, V = 20),
                  Cp2 = calc_ss_1cmt_linear_infusion(t = t, tau = 24, dose = 600, CL = 1, V = 20, tinf=2),
                  Cp3 = calc_ss_1cmt_linear_oral_0(t = t, tau = 24, dose = 600, CL = 1, V = 20, dur=2),
                  Cp4 = calc_ss_1cmt_linear_oral_0_lag(t = t, tau = 24, dose = 600, CL = 1, V = 20, dur=2, tlag=1.5),
                  Cp5 = calc_ss_1cmt_linear_oral_1(t = t, tau = 24, dose = 600, CL = 1, V = 20, ka=1.5),
                  Cp6 = calc_ss_1cmt_linear_oral_1_lag(t = t, tau = 24, dose = 600, CL = 1, V = 20, ka=1.5, tlag=1.5))

p1.1ss <- ggplot(df1ss, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p1.2ss <- ggplot(df1ss, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p1.3ss <- ggplot(df1ss, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p1.4ss <- ggplot(df1ss, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p1.5ss <- ggplot(df1ss, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p1.6ss <- ggplot(df1ss, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p1.1ss, p1.2ss, p1.3ss, p1.4ss, p1.5ss, p1.6ss, nrow=3)

```

## Two-compartment, single dose 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
2-compartment   | Single-dose, bolus                                     | `calc_sd_2cmt_linear_bolus(t, dose, CL, V1, V2, Q)`
&nbsp;          | Single-dose, infusion                                  | `calc_sd_2cmt_linear_infusion(t, dose, CL, V1, V2, Q, tinf)` 
&nbsp;          | Single-dose, zero-order oral absorption                | `calc_sd_2cmt_linear_oral_0(t, dose, CL, V1, V2, Q, dur)`
&nbsp;          | Single-dose, zero-order oral absorption with lag time  | `calc_sd_2cmt_linear_oral_0_lag(t, dose, CL, V1, V2, Q, dur, tlag)`
&nbsp;          | Single-dose, first-order oral absorption               | `calc_sd_2cmt_linear_oral_1(t, dose, CL, V1, V2, Q, ka)` 
&nbsp;          | Single-dose, first-order oral absorption with lag time | `calc_sd_2cmt_linear_oral_1_lag(t, dose, CL, V1, V2, Q, ka, tlag)` 

For example -

```{r 2cptdemo, fig.height=10, fig.width=7, message=FALSE}

t <- seq(0, 24, by=0.1)

df2 <- data.frame(t = t,
                  Cp1 = calc_sd_2cmt_linear_bolus(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1),
                  Cp2 = calc_sd_2cmt_linear_infusion(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1, tinf=2),
                  Cp3 = calc_sd_2cmt_linear_oral_0(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1, dur=2),
                  Cp4 = calc_sd_2cmt_linear_oral_0_lag(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1, dur=2, tlag=1),
                  Cp5 = calc_sd_2cmt_linear_oral_1(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1, ka=1.5),
                  Cp6 = calc_sd_2cmt_linear_oral_1_lag(t = t, dose = 600, CL = 5, V1 = 20, V2 = 80, Q = 1, ka=1.5, tlag=1))

p2.1 <- ggplot(df2, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p2.2 <- ggplot(df2, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p2.3 <- ggplot(df2, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p2.4 <- ggplot(df2, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p2.5 <- ggplot(df2, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p2.6 <- ggplot(df2, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p2.1, p2.2, p2.3, p2.4, p2.5, p2.6, nrow=3)

```

## Two-compartment, steady state 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
2-compartment   | Steady state, bolus                                     | `calc_ss_2cmt_linear_bolus(tad, tau, dose, CL, V1, V2, Q)`
&nbsp;          | Steady state, infusion                                  | `calc_ss_2cmt_linear_infusion(tad, tau, dose, CL, V1, V2, Q, tinf)` 
&nbsp;          | Steady state, zero-order oral absorption                | `calc_ss_2cmt_linear_oral_0(tad, tau, dose, CL, V1, V2, Q, dur)`
&nbsp;          | Steady state, zero-order oral absorption with lag time  | `calc_ss_2cmt_linear_oral_0_lag(tad, tau, dose, CL, V1, V2, Q, dur, tlag)`
&nbsp;          | Steady state, first-order oral absorption               | `calc_ss_2cmt_linear_oral_1(tad, tau, dose, CL, V1, V2, Q, ka)` 
&nbsp;          | Steady state, first-order oral absorption with lag time | `calc_ss_2cmt_linear_oral_1_lag(tad, tau, dose, CL, V1, V2, Q, ka, tlag)` 

For example -

```{r 2cptssdemo, fig.height=10, fig.width=7,message=FALSE}

t <- seq(0, 24, by=0.1)

df2ss <- data.frame(t = t,
                  Cp1 = calc_ss_2cmt_linear_bolus(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25),
                  Cp2 = calc_ss_2cmt_linear_infusion(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25, tinf=2),
                  Cp3 = calc_ss_2cmt_linear_oral_0(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25, dur=2),
                  Cp4 = calc_ss_2cmt_linear_oral_0_lag(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25, dur=2, tlag=1.5),
                  Cp5 = calc_ss_2cmt_linear_oral_1(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25, ka=1.5),
                  Cp6 = calc_ss_2cmt_linear_oral_1_lag(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 80, Q = 0.25, ka=1.5, tlag=1.5))

p2.1ss <- ggplot(df2ss, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p2.2ss <- ggplot(df2ss, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p2.3ss <- ggplot(df2ss, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p2.4ss <- ggplot(df2ss, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p2.5ss <- ggplot(df2ss, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p2.6ss <- ggplot(df2ss, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p2.1ss, p2.2ss, p2.3ss, p2.4ss, p2.5ss, p2.6ss, nrow=3)

```

## Three-compartment, single dose 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
3-compartment   | Single-dose, bolus                                     | `calc_sd_3cmt_linear_bolus(t, dose, CL, V1, V2, V3, Q2, Q3)`
&nbsp;          | Single-dose, infusion                                  | `calc_sd_3cmt_linear_infusion(t, dose, CL, V1, V2, V3, Q2, Q3, tinf)` 
&nbsp;          | Single-dose, zero-order oral absorption                | `calc_sd_3cmt_linear_oral_0(t, dose, CL, V1, V2, V3, Q2, Q3, dur)`
&nbsp;          | Single-dose, zero-order oral absorption with lag time  | `calc_sd_3cmt_linear_oral_0_lag(t, dose, CL, V1, V2, V3, Q2, Q3, dur, tlag)`
&nbsp;          | Single-dose, first-order oral absorption               | `calc_sd_3cmt_linear_oral_1(t, dose, CL, V1, V2, V3, Q2, Q3, ka)` 
&nbsp;          | Single-dose, first-order oral absorption with lag time | `calc_sd_3cmt_linear_oral_1_lag(t, dose, CL, V1, V2, V3, Q2, Q3, ka, tlag)` 

For example -

```{r 3cptdemo, fig.height=10, fig.width=7, message=FALSE}

t <- seq(0, 24, by=0.1)

df3 <- data.frame(t = t,
                  Cp1 = calc_sd_3cmt_linear_bolus(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600),
                  Cp2 = calc_sd_3cmt_linear_infusion(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600, tinf=2),
                  Cp3 = calc_sd_3cmt_linear_oral_0(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600, dur=2),
                  Cp4 = calc_sd_3cmt_linear_oral_0_lag(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600, dur=2, tlag=1),
                  Cp5 = calc_sd_3cmt_linear_oral_1(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600, ka=1.5),
                  Cp6 = calc_sd_3cmt_linear_oral_1_lag(t = t, CL = 3.5, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dose = 600, ka=1.5, tlag=1))

p3.1 <- ggplot(df3, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p3.2 <- ggplot(df3, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p3.3 <- ggplot(df3, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p3.4 <- ggplot(df3, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p3.5 <- ggplot(df3, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p3.6 <- ggplot(df3, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p3.1, p3.2, p3.3, p3.4, p3.5, p3.6, nrow=3)

```

## Three-compartment, steady state 

Available functions include:

Model                         | Variant                                    | Call              
------------------------------|--------------------------------------------|----------------------------------------
3-compartment   | Single-dose, bolus                                     | `calc_ss_3cmt_linear_bolus(tad, tau, dose, CL, V1, V2, V3, Q2, Q3)`
&nbsp;          | Single-dose, infusion                                  | `calc_ss_3cmt_linear_infusion(tad, tau, dose, CL, V1, V2, V3, Q2, Q3, tinf)` 
&nbsp;          | Single-dose, zero-order oral absorption                | `calc_ss_3cmt_linear_oral_0(tad, tau, dose, CL, V1, V2, V3, Q2, Q3, dur)`
&nbsp;          | Single-dose, zero-order oral absorption with lag time  | `calc_ss_3cmt_linear_oral_0_lag(tad, tau, dose, CL, V1, V2, V3, Q2, Q3, dur, tlag)`
&nbsp;          | Single-dose, first-order oral absorption               | `calc_ss_3cmt_linear_oral_1(tad, tau, dose, CL, V1, V2, V3, Q2, Q3, ka)` 
&nbsp;          | Single-dose, first-order oral absorption with lag time | `calc_ss_3cmt_linear_oral_1_lag(tad, tau, dose, CL, V1, V2, V3, Q2, Q3, ka, tlag)`

For example -

```{r 3cptssdemo, fig.height=10, fig.width=7,message=FALSE}

t <- seq(0, 24, by=0.1)

df3ss <- data.frame(t = t,
                  Cp1 = calc_ss_3cmt_linear_bolus(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05),
                  Cp2 = calc_ss_3cmt_linear_infusion(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, tinf=2),
                  Cp3 = calc_ss_3cmt_linear_oral_0(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dur=2),
                  Cp4 = calc_ss_3cmt_linear_oral_0_lag(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, dur=2, tlag=1.5),
                  Cp5 = calc_ss_3cmt_linear_oral_1(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, ka=1.5),
                  Cp6 = calc_ss_3cmt_linear_oral_1_lag(tad = t, tau = 24, dose = 600, CL = 1, V1 = 20, V2 = 500, V3 = 200, Q2 = 0.5, Q3 = 0.05, ka=1.5, tlag=1.5))

p3.1ss <- ggplot(df3ss, aes(t, Cp1)) +
  geom_line() +
  labs(title="Bolus dose")

p3.2ss <- ggplot(df3ss, aes(t, Cp2)) +
  geom_line() +
  labs(title="Infusion")

p3.3ss <- ggplot(df3ss, aes(t, Cp3)) +
  geom_line() +
  labs(title="Zero-order oral")

p3.4ss <- ggplot(df3ss, aes(t, Cp4)) +
  geom_line() +
  labs(title="Zero-order oral with lag time")

p3.5ss <- ggplot(df3ss, aes(t, Cp5)) +
  geom_line() +
  labs(title="First-order oral")

p3.6ss <- ggplot(df3ss, aes(t, Cp6)) +
  geom_line() +
  labs(title="First-order oral with lag time")

grid.arrange(p3.1ss, p3.2ss, p3.3ss, p3.4ss, p3.5ss, p3.6ss, nrow=3)

```

## Multiple dosing

Being able to display single-dose and steady state curves for various commonly-seen systems is all very well but it is often more useful to show a concentration-time curve which demonstrates accumulation and time to steady state. The `pk_curve()` function makes this straightforward by using the principle of superposition.

```{r pkcurve1, fig.height=4, fig.width=7,message=FALSE}

dfcurve1 <- pk_curve(t=seq(0,168,by=0.1), model="3cmt_oral", ii=24, addl=12,
                     params=list(CL=1.5, V1=25, V2=2, V3=5, Q2=0.5, Q3=0.25, ka=1))

ggplot(dfcurve1, aes(t, cp)) +
  geom_line()

```







                